Notes on conformal manifolds

### 2025-09-23 Draft_MLFlatDirections.tex
Updates of AIS-SMC section with Bastien.

%!TEX output_directory = aux

@@

@@

	We initialized the algorithm with
	\begin{equation}	\label{eq:sample10000adaptativetemp}
		\begin{aligned}
			&\texttt{data\_size = 10,000},\\
			&\texttt{n\_epochs = 1000},\\
			&\texttt{beta  = 1e-6}, \\
			&\texttt{p\_{modify} = 0.5},\\
			&\texttt{p\_{multiply} = 0.25},\\
			&\texttt{p\_{divide} = 0.25},\\
			&\texttt{lambda = 1000},\\
			&\texttt{sigma = 0.1},\\
			&\texttt{n\_particles = 1000}.
		\end{aligned}
	\end{equation}
	%ICI \gl{link to notations in section 3}
	The convenience of these parameters was determined empirically --a systematic analysis of its optimization is left for future work. For the representation of polynomials, we adopt a vectorial approach: each polynomial is represented as an array of its coefficients $\{c_k\}$, where $k \in \{1,\dots,n_{\text{mon}}\}$ and $n_{\text{mon}}$ denotes the number of available monomials, determined by the chosen maximum degree \texttt{max\_degree} and the number of variables. We also impose a constraint on the maximum number of terms within a single polynomial \texttt{max\_num\_monomials} to promote sparsity. \footnote{An alternative approach would be to encourage sparsity through the prior distribution or within the loss function formulation.} For our simulations, we chose:
	\begin{equation} \label{eq:parampol}
		\begin{aligned}
			&\texttt{max\_degree = 4},\\
			&\texttt{max\_num\_monomials = 6.}
		\end{aligned}
	\end{equation}
With these parameters and given that we are dealing with $5$ variables, $n_{\text{mon}} = 126$.

	Regarding the temperature schedule, as can be seen in \eqref{eq:gamman}, what matters to determine $\gamma_{n}$ is not the inverse temperature, but rather the product of the inverse temperature and the loss function. Therefore, instead of implementing an arbitrary schedule for $\beta$, we employed an adaptive temperature approach that sets the number of particles to be updated at each run as
		\begin{equation} \label{eq:eqadaptivetemperatureschedule}
			\texttt{adaptive\_temperature\_ratio} = 0.8 - 0.5 \times \left(\frac{\texttt{epoch}}{\texttt{n\_epochs}}\right).
		\end{equation}
This encourages exploration during the early stages of training and gradually transitions to a more selective regime. %ICI \bd{Move \eqref{eq:eqadaptivetemperatureschedule} in section 4?}

>>>>

@@

	\subsection{General idea}
	The numerical sampling of the space of vacua described above results in clouds of points embedded in a higher-dimensional space. 
	The aim of symbolic regression is to uncover interpretable mathematical expressions that best describe the embedding of these loci of solutions. 
	In the present context, the form of the supergravity potential \eqref{eq:scalarpotential} determining the solutions implies that the loci can be defined through polynomials $p_{m}$ such that $p_{m}(\vec{x}^{(i)}) = 0$, with $i \in \{1, \dots, n_{\rm points}\}$. Such polynomials will be called ``annihilating polynomials'' in the following.
	Finding annihilating polynomials involves navigating a vast, discrete, and often rugged search space of possible symbolic models, which poses significant challenges for traditional sampling methods. 
	Markov Chain Monte Carlo (MCMC) techniques such as Metropolis-Hastings or Gibbs sampling, while widely used, can struggle with poor mixing and often get trapped in local optima, especially in high-dimensional or multimodal spaces \cite{??}. \ce{Citation.}

	To address these challenges, we formulate the search for annihilating polynomials as a determination of a density probability on the space of polynomials that stresses higher probabilities on polynomials $p$ that minimises $\displaystyle \sum_{i=1}^{n_{\rm points}}p(\vec{x}^{(i)})^{2}$. Such a density probability is \textit{a priori} hard to find, but can be approximated using variations of Important Sampling methods~\cite{910e8e7d-2e70-317c-adf6-e21cd50755a1}. Importance sampling is a method to approximate expectation values with respect to a target distribution $d$ from a weighted average from another distribution $\pi$. It is based on the observation that the expectation
	\begin{equation}
		\mathbb{E}_{d}[f] = \int f(z)d(z)\ \d z
	\end{equation}
	of a function $f$ with respect to $d$ can be computed from the expectation $\mathbb{E}_{\pi}[w f]$, now computed with respect to the distribution $\pi$, with unnormalised $w(z) = d(z)/\pi(z)$. Therefore, if it is easier to draw a sample $\{z^{(k)}\}_{k\in\llbracket 1, n_{\rm particles}\rrbracket}$ from $\pi$ than from $d$, one can estimate $\mathbb{E}_{d}[f]$ from the weighted average
	\begin{equation}
		\mathbb{E}_{d}[f] \simeq \frac{\displaystyle \sum_{k=1}^{n_{\rm particles}} w(z^{(k)})\,f(z^{(k)})}{\displaystyle \sum_{i=1}^{n_{\rm particles}} w(z^{(k)})}.
	\end{equation}
	The accuracy of this estimation depends on the size of the sample and, crucially, on the choice of the distribution $\pi$.

	To have an accurate estimation of $d$, we employ an Annealed Importance Sampling (AIS) \cite{neal1998annealedimportancesampling} combined with Sequential Monte Carlo (SMC) methods \cite{del2006sequential,naesseth2024elementssequentialmontecarlo}. AIS aims at estimating $d$ sequentially from multiple intermediate distributions $\pi_{n}$ that smoothly transition from an initial, tractable distribution $d_{0}$ (chosen to be easy to sample from) to the complex target distribution $d$.
	These distributions are approximated by weighted samples $\{z^{(k)}_{n},w^{(k)}_{n}\}$ of particles, each representing a candidate symbolic expression, such that the resulting weighted average estimates $\pi_{n}$: \ce{Normalise the weights?}
	\begin{equation}
		\frac{\displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}_{n}\,f(z^{(k)}_{n})}{\displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}_{n}} \underset{n_{\rm particles}\rightarrow+\infty}{\longrightarrow} \mathbb{E}_{\pi_{n}}[f],
	\end{equation}
	or equivalently
	\begin{equation} \label{eq:pifromweights}
		\pi_{n}(z) \simeq \frac{\displaystyle \sum_{k=1}^{N} w^{(k)}_{n}\,\delta(z-z^{(k)}_{n})}{\displaystyle \sum_{k=1}^{N} w^{(k)}_{n}}.
	\end{equation}
	The $\pi_{n}$ are incrementally deformed using a Markov kernel to gradually reproduce the target $d$ and the particles get reweighed. This annealing process is guided by a temperature-like parameter that gradually emphasises the data likelihood, allowing for more efficient and tunable exploration of the polynomials landscape.

	This procedure has the disadvantage that the weights variance tends to increase, leading to weight degeneracy. We use SMC methods to incorporate a resampling procedure to AIS, thus focussing computational effort on high-probability regions.  This combination of importance sampling, mutation and resampling maintains diversity among the particles and prevents premature convergence to suboptimal models. These features make AIS-SMC particularly well-suited for symbolic regression tasks, where the search space is not only high-dimensional but also structured and discontinuous.

	\subsection{Detailed procedure}

	Let us now explain in more detail the procedure. 
	The goal is to obtain a probability distribution $d(z)$, where $z$ runs over a space of polynomials, that assigns high probabilities to those polynomials that annihilate the data, and therefore will be used to identify their symbolic expressions.
	We reconstruct this density function $d$ by series of density functions $\pi_n(z)$, $n\in\llbracket1,n_{\rm epochs}\rrbracket$, such that $\pi_n\to d$ as $n\to n_{\rm epochs}$. 
	On every annealing step, $\pi_n$ is defined in terms of an unnormalised density $\gamma_n$ and a normalisation constant as $\pi_n(z) = \gamma_n(z)/Z_n$.
	The unnormalised density at level $n$ is given in terms of a prior distribution $d_0(z)$ over the space of polynomials and a loss function $L(z)$, as \ce{Citation.}
			\begin{equation}\label{eq:gamman}
				\gamma_n(z) = d_0(z) \: \exp \Big( -\beta_n L(z)\Big),
			\end{equation} 
	where the inverse temperature constants $\beta_n$ are taken to evolve monotonically as $\beta_1 < \beta_2 <\dots< \beta_{n_{\rm epochs}}$. The importance given to each polynomial $z$ is thus updated at each steps, with polynomials $z$ with high loss being disfavoured. This update is controlled by the inverse temperature constants, enabling deeper exploration during the early stages and protecting the procedure from being trapped in local minima. The loss function is chosen to be minimal on annihilating polynomials, its design will be discussed at the end of the section. We chose the prior distribution $d_{0}$ to be flat on the space of polynomials, giving equal importance to all of them, which makes easy the initial sampling of the particles $\{z^{(k)}_{0}\}$.

	At each epoch $n$, $\pi_{n}$ is approximated \textit{via} a set of particles and weights $\{z_{n}^{(k)},\ w_{n}^{(k)}\}$ in the sense of eq.~\eqref{eq:pifromweights}. The sample of particles $z^{(k)}_{n}$ get mutated using a forward propagation Markov kernel $q(z^{(k)}_{n+1}\vert z^{(k)}_{n})$, defining the probability to get $z^{(k)}_{n+1}$ when the current state is $z^{(k)}_{n}$. For the implementation of the forward propagation kernel, we perform an AIS-style MCMC algorithm. 
	On every epoch, we make some move in the space of polynomials, and then accept or reject those new polynomials based on a fixed rate.
	The moves that have been allowed in our implementation are the following.
	\begin{itemize}
		\item \textbf{Coefficient pertubation}\quad Given a polynomial, we choose one of its coefficients at random and modify it by a Gaussian noise distributed as $\mathcal{N}(0,\sigma^2)$, \textit{e.g.}
		\begin{equation}
			2 x_1x_{2} + 3 x_2^2 \longmapsto  2.1 x_1x_{2} + 3 x_2^2\,.
		\end{equation}
		\item \textbf{Variable multiplication}\quad Given a polynomial, we pick randomly one of its monomials, and multiply it by one of the available variables, \textit{e.g.}
		\begin{equation}
			2 x_1x_{2} + 3 x_2^2 \longmapsto  2 x_1 x_2^{2} + 3 x_2^2\,.
		\end{equation}
		\item \textbf{Variable division}\quad Given a polynomial, we choose randomly one of its monomials, and divide it by one of its variables, \textit{e.g.}
		\begin{equation}
			2 x_1 x_2 + 3 x_2^2 \longmapsto  2 x_1 + 3 x_2^2\,.
		\end{equation}
	\end{itemize}
	Each of these operations is chosen randomly, with probabilities $p_{\rm shift}$, $p_{\rm multiply}$, and $p_{\rm divide}$.
	%		For the case of coefficient perturbations, the transition is symmetric. 
	%		However, this symmetry does not hold in general. For other types of operations—such as multiplication or division by a variable—the transition from \( z_{n-1}^k \) to \( z_n^k \) is generally not equivalent to the reverse transition from \( z_n^k \) to \( z_{n-1}^k \). In such cases $q(z_n^k \mid z_{n-1}^k) \ne q(z_{n-1}^k \mid z_n^k)$ and particular care must be taken in the computation of the backward propagation kernel.
	After performing these updates, the change is accepted based on the temperature-dependent rate
	\begin{equation}\label{eq:acceptanceratio}
		A(z_{n+1}^{(k)},z_{n}^{(k)}) = \mathrm{min} \left(1,\frac{ \gamma_{n+1}(z_{n+1}^{(k)}) q(z_{n}^{(k)} | z_{n+1}^{(k)}) }{ \gamma_{n}(z_{n}^{(k)}) q(z_{n+1}^{(k)} | z_{n}^{(k)}) }  \right)\,.
	\end{equation}
	We then draw $u\sim \mathrm{Uniform}(0,1)$, accept the new particle if $u<A(z_{n+1}^{(k)},z_{n}^{(k)})$ and reject it otherwise.
	The kernel $q(z_{n-1}^k | z_n^k)$ thus encapsulates the effects of performing these operations on the polynomials at step $n$. \ce{Why this comment?}

	Once the particles have been mutated, the evolution of the weights is given by
	\begin{equation}
		w_{n+1}^{(k)} = w_{n}^{(k)}\ \alpha_{n+1}^{(k)},
	\end{equation}
	where the incremental importance weights $\alpha_{n+1}^{(k)}$ are given by\footnote{The incremental weight is in general defined as $\alpha^{(k)}_{n+1} = \frac{\gamma_{n+1}(z_{n+1}^{(k)})}{\gamma_{n}(z_n^{(k)})}\,\frac{K(z^{(k)}_{n}\vert z^{(k)}_{n+1})}{q(z^{(k)}_{n+1}\vert z^{(k)}_{n})}$, where $K(z^{(k)}_{n}\vert z^{(k)}_{n+1})$ is the backward propagation kernel~\cite{del2006sequential}. For simplicity, we chose to identify $K$ to the forward propagation kernel $q$.}~\cite{del2006sequential}
	\begin{equation}	\label{eq: incremental_importance}
			\alpha^{(k)}_{n+1} = \frac{\gamma_{n+1}(z_{n+1}^{(k)})}{\gamma_{n}(z_n^{(k)})}\,\frac{q(z^{(k)}_{n}\vert z^{(k)}_{n+1})}{q(z^{(k)}_{n+1}\vert z^{(k)}_{n})}=\frac{\exp\big(-\beta_{n+1}L(z_{n+1}^{(k)})\big)}{\exp\big(-\beta_{n}L(z_{n}^{(k)})\big)}\,\frac{q(z^{(k)}_{n}\vert z^{(k)}_{n+1})}{q(z^{(k)}_{n+1}\vert z^{(k)}_{n})}.
	\end{equation}

	To ensure an efficient exploration of the landscape, we would like to avoid having to many particles with low weights: as the algorithm progresses, particles with negligible weights contribute minimally to the approximation of the target distribution (see eq.~\eqref{eq:pifromweights}), leading to a concentration of the effective sample on a subset of particles with substantial weights. This can be estimated using the effective sample size~\cite{Liu01091998}
	\begin{equation}
		{\rm ESS}_{n+1} = \Bigg(\sum_{k=1}^{n_{\rm particles}} \tilde{w}^{(k)}_{n+1}{}^{2}\Bigg)^{-1},
	\end{equation}
	where the $\tilde{w}^{(k)}_{n+1}$ are normalised weights deduced from the $w^{(k)}_{n+1}$. ${\rm ESS}_{n+1}$ takes values between $1$ (complete degeneracy, there is only one meaningful particle) and $n_{\rm particles}$ (no degeneracy, all particles contribute equally). We resample the particles when the ESS falls below a predetermined threshold, typically $n_{\rm particles}/2$. During resampling, new particles are drawn from the empirical distribution defined by the current particle weights $\tilde{w}^{(k)}_{n+1}$.	This procedure results in a new particle set where high-weight particles may appear multiple times, while low-weight particles may be eliminated entirely. Following resampling, all particle weights are reset to uniform values $w_{n+1}^{(k)} = 1/n_{\rm particles}$, as the information previously encoded in the weight distribution has been incorporated into the spatial distribution of the resampled particles. This resampling mechanism ensures that computational resources are concentrated on exploring the most promising regions of the state space.

	A flow chart of the full AIS-SMC procedure is given in fig.~\ref{ASMCflowchart}. Note that in practice the reweighing and resampling of particles $z_{n-1}^{(k)}$ are done at the beginning of the epoch $n$.

	\paragraph{}
	To compute the acceptance ratio and weights \textit{via} eq.~\eqref{eq:acceptanceratio} and \eqref{eq: incremental_importance} and we need to consider the loss function in~\eqref{eq:gamman}.
	In the present context, we use the following loss, where the polynomials have been parameterised as $z = \sum_m c_m X_m$ with $X_m$ denoting the possible monomials up to a given degree,%
	\footnote{So, if we have $x$ and $y$ as variables, and the maximum degree is 2, then $X_m$ are $1, x, y, x^2, y^2, x y$.}
	\begin{equation} \label{eq:lossreg}
		L(z) = \sum_{i=1}^{n_{\rm points}} z(x^{(i)})^2 + \frac{\lambda}{\sum_m |c_m|}\,.
	\end{equation}
	Here, the first term is just the sum of the square of the polynomial evaluated on the data, and is therefore minimal when the polynomial annihilates the data. 
	The second term is a regularisation factor that prevents the algorithm to send all the coefficient to 0, which would give a trivial solution to the problem. 
	We typically take $\lambda \sim \mathcal{O}(10^3)$. 
	Together with the loss function, we also need to specify a cooling schedule for $\beta$. We will discuss different approaches in the next section.

	\paragraph{}
	Once the Annealing loop is over, we end up with a total of $n_{\rm particles}$ particles, which in principle should be close to annihilate our data, but whose coefficients may need some refinement. To deal with it, for each polynomial we run a fine tuning loop as described in sec.~\ref{sec:1runanalysis}.

@@

	We initialized the algorithm with
	\begin{equation}	\label{eq:sample10000adaptativetemp}
		\begin{aligned}
			&\texttt{n\_points = 10,000},\\
			&\texttt{n\_epochs = 1000},\\
			&\texttt{beta  = 1e-6}, \\
			&\texttt{p\_{modify} = 0.5},\\
			&\texttt{p\_{multiply} = 0.25},\\
			&\texttt{p\_{divide} = 0.25},\\
			&\texttt{lambda = 1000},\\
			&\texttt{sigma = 0.1},\\
			&\texttt{n\_particles = 1000}.
		\end{aligned}
	\end{equation}
	%ICI \gl{link to notations in section 3}
	The convenience of these parameters was determined empirically --a systematic analysis of its optimization is left for future work. For the representation of polynomials, we adopt a vectorial approach: each polynomial is represented as an array of its coefficients $\{c_m\}$, where $m \in \{1,\dots,n_{\text{mon}}\}$ and $n_{\text{mon}}$ denotes the number of available monomials, determined by the chosen maximum degree \texttt{max\_degree} and the number of variables. We also impose a constraint on the maximum number of terms within a single polynomial \texttt{max\_num\_monomials} to promote sparsity. \footnote{An alternative approach would be to encourage sparsity through the prior distribution or within the loss function formulation.} For our simulations, we chose:
	\begin{equation} \label{eq:parampol}
		\begin{aligned}
			&\texttt{max\_degree = 4},\\
			&\texttt{max\_num\_monomials = 6.}
		\end{aligned}
	\end{equation}
With these parameters and given that we are dealing with $5$ variables, $n_{\text{mon}} = 126$.

Regarding the temperature schedule, as can be seen in \eqref{eq:gamman}, what matters to determine $\gamma_{n}$ is not the inverse temperature, but rather the product of the inverse temperature and the loss function. Therefore, instead of implementing an arbitrary schedule for $\beta$, we employed an adaptive temperature approach, whose aim is to ensure that the number of particles to be updated at each run is close to
\begin{equation} \label{eq:eqadaptivetemperatureschedule}
	\texttt{adaptive\_temperature\_ratio} = 0.8 - 0.5 \times \left(\frac{\texttt{n}}{\texttt{n\_epochs}}\right).
\end{equation}
In other words, if the number of accepted particles after the MCMC step at a given \texttt{n} is higher than the target~\eqref{eq:eqadaptivetemperatureschedule}, we decrease the temperature and hence increase $\beta_{n}$, and \textit{vice versa} if this number is lower than \texttt{adaptive\_temperature\_ratio}.

This encourages exploration during the early stages of training and gradually transitions to a more selective regime.
